---
title: "Module 3.1: Descriptive Statistics for a single variable"
author: "David Quigley"
date: "September 25, 2015"
output: 
    html_document:
        toc: true
        css: swiss.css
---


```{r global_options, include=FALSE}
knitr::opts_chunk$set(comment = NA)
knitr::opts_chunk$set( prompt=TRUE ) 

knitr::knit_hooks$set(small.mar = function(before, options, envir) {
    if (before) par(mar = c(2, 2, 1, 1))  # smaller margin on top and right
})
knitr::knit_hooks$set(left.mar = function(before, options, envir) {
    if (before) par(mar = c(2, 4, 1, 1))  # smaller margin on top and right
})
knitr::knit_hooks$set(med.mar = function(before, options, envir) {
    if (before) par(mar = c(4, 4, 1, 1))  # smaller margin on top and right
})
knitr::knit_hooks$set(bottom.mar = function(before, options, envir) {
    if (before) par(mar = c(3, 2, 1, 1))  # smaller margin on top and right
})
```


```{r echo=FALSE}
height = c(7,8,7,9.5)
weight = c(4,3.2, 3.7, 5.4)
myTable=data.frame(height, weight)
rownames(myTable)=c('Flopsy','Mopsy','Cottontail','Peter')

my_observations = rnorm(1000, mean=50, sd=10)
my_bivariate = c( rnorm(1000, mean=50, sd=5), rnorm(1000, mean=65, sd=5) )
pap_no = c( 9,17,9,0,15,12,7,0,0,0,13,10,0,0,0,8,1,7,4,
  7,2,12,1,4,0,35,6,9,0,0,12,16,16,0,6,3,0,4,
  0,0,0,0,3,2,1,7,5,23,2,3,3,5,0,3,5,17,1,
  1,4,21,0,0,2,14,5,2,0,3,6,2,1)

MCV = c( 59.5,49.0,48.7,50.1,49.9,54.5,57.0,63.2,47.5,65.1,63.5,61.3,77.3,50.2,50.9,66.4,62.1,49.7,58.2,60.0,54.8,45.3,48.8,61.2,47.8,53.4,60.1,61.4,
57.9,61.2,62.1,64.6,61.3,45.8,65.9,47.6,59.4,63.5,52.2,51.8,75.9,51.3,60.6,55.7,69.3,50.5,57.7,54.6,67.7,60.5,54.4,58.6,73.6,60.0,53.2,49.6,
56.4,62.0,48.9,52.9,50.1,56.3,56.6,64.3,59.3,57.3,57.2,48.1,46.6,45.2,48.7,60.6,48.0,47.5,52.7,49.7,56.4,51.2,46.1,43.8,51.7,47.4,59.0,54.1,
47.9,43.6,47.6,53.7,45.2,48.7,49.7,46.6,51.2,46.1,50.6,54.2,47.4,52.7,51.7,50.6,49.5,55.1,51.9,59.6,46.3,51.0,48.3,50.2,56.5,56.3,46.5,45.8,
55.3,47.0,49.8,47.6,56.9,55.4,52.9,58.1,58.2,56.0,57.0,48.6,55.4,58.3,57.2,58.6,55.0,65.5,60.7,45.5,44.1,47.2,50.4,46.7,55.1,52.5,53.1,55.2,
52.3,50.5,48.9,45.2,53.3,48.4,55.0,55.4,49.2,43.6,52.2,51.6,49.0,51.0,60.4,45.7,55.0,50.5,46.8,46.4,46.5,54.4,46.1,48.4,56.1,55.2,56.8,58.1,
47.7,55.0,58.6,55.2,47.2,46.5,58.2,56.7,58.9,57.3,54.2,54.6,54.9,55.7,54.7,56.1,55.3,51.1,48.3,46.2,51.6,54.7,49.9,54.0,55.8,49.6,46.2,58.4,
52.4,44.7,47.8,59.2,57.3,49.6,45.9,44.3,57.0,44.4,47.8,49.4,50.7,56.0,47.6,52.8,46.0,47.6,51.4,55.4,45.5,46.2,47.1,51.3,52.4,49.4,48.0,46.2)

```

Introduction
========================================


**This module explains:**

* The definition of a random variable
* How to visualize the distribution of a single variable
    + scatter plots
    + bar plots
    + histograms



Types of data
===============

Experiments observe the outcome of random processes
------------------------------------------------------------

Experiments, whether they are performed in the laboratory or in the field, are designed to collect data. These data can take many forms: counts: binary yes/no values indicating whether a person has had a heart attack, numbers indicating blood pressure readings, categories indicating type of brain which type of dementia patients have experienced. Statisticians call these observed data *random variables* to indicate that their values were not pre-determined. The use of the word "random" does *not* mean that the observations are entirely the result of chance, like flipping a coin. It means that the value for a given observation is due, in part, to chance effects, and repetition of the experiment would not produce identical results. We use the term *random variable* even for data where all of the possible values for that observation can be enumerated. 

Calling a variable "random" is *not* the same as saying we have no ability to predict its likely value.

Variables have qualities that allow us to place them into one of several categories. These qualities are important because they determine what kind of analysis can be performed. The most frequently encountered categories of observation are *continuous*, *ordinal*, and *categorical* data.

Continuous variables
----------------------------

Continuous data are numbers that can take any real value; that is, they could be anywhere on a number line. Height and weight are continuous variables. Age in days is continuous, even if we constrain it to be rounded to the nearest whole number. Numbers that are constrained to be integers are a special case of continuous variables. 

Ordinal variables
----------------------------

Variables that are not numbers, but which can be sorted into an invariant order, are ordinal. Ordinal variables do not need to have a specific numerical range. "weight group: Low, Medium, High" and "tumor grade: Grade I, Grade IIA, Grade IIB, Grade III, Grade IV" are ordinal variables. Some specialized statistics such as the Cochran–Armitage test for trend can be applied to ordinal variables.

Categorical variables
----------------------------

Categorical data are observations that fall into cleanly definable groups that lack an ordering. The eye color of a patient can be coded as a categorical variable: (*Brown*, *Blue*, *Green*, *Grey*).  

Categorical variables and Factors
--------------------------------------------------------

Statisticians often refer to categorical variables as *factors*. Although categorical variables are sometimes coded using numbers (e.g. "red=1, blue=2, orange=3...", this can lead to confusion if you apply a statistical test that interprets the numbers as having an inherent meaning. R has a function *factor()* which converts a vector of categorical variables, usually words, into factors. The variable then retains its appearance as words, but R will know that these words should be treated as categories.

Counts are derived from categorical data
------------------------------------------

Counts are integers of value zero or higher. Counts refer to the number of times we observed the outcome of some process. As such, counts are not a primary data type, but rather a summary of the number of times we saw each possible element in a categorical or ordinal group. Counts are of particular interest because they are the appropriate data for Fisher's Exact Test and the chi-squared test.


The same data can be represented by different types of variables
================================================================================

**The form of a variable involves a choice by the investigator**. This choice has consequences. As an example, consider Body Mass Index (BMI), a measure of obesity that incorporates height and weight. BMI is a continuous variable. You could make different choices about how to represent BMI in your experiment.

* **Continuous variable**: report the raw BMI value. 
* **Ordinal variable**: divide subjects into Low, Medium, and High BMI. To place subjects into a category requires making a decision about the cut-off values for each category. Different choices of cut-off can have large consequences for the interpretation of the data you collect.
* **Categorical variable**: divide patients into Obese and Not Obese categories. Again, the choice of cut-off is crucial. 

continuous | ordinal | categorical
---------- | ------- | -----------
25         | medium  | not-obese
21         | low     | not-obese
41         | high    | obese
32         | high    | obese
29         | medium  | not-obese
42         | high    | obese


**The same BMI data, three ways**

The choice of how to fit continuous data into categories is very important. In some analysis, where there is no common convention, you will be able to decide on your own categories. These decisions have to be well-motivated, and ideally should take place before you see the data that will be generated. If the categories are fit to the data *post hoc*, they can be a serious source of bias.

In some cases, standardized value for the ranges of data that should fit into a particular category will be established by informal common scientific conventions or by formal standards bodies. Knowing whether these standard definitions apply to your analysis is extremely important for clear communication and for clinical relevance if your study describes medical information. 


Displaying observations of a single variable
================================================

Data should be communicated in as informative and straightforward a manner as possible. The human eye is one of the most important tools in statistical analysis. Statistical analysis is partly a mathematical exercise, but it is also an exploratory exercise. Always look at the data you are trying to interpret. 

The choice of visual plot is dictated in part by the form of the data, in part by the specific question you wish to ask of the data, and in part by the conventions and expectations of the community of scientists to whom you want to communicate. 

Scatter plots and Bar plots
------------------------------------

One choice for continuous data is to make a one-dimensional plot of the values. As an example I'd like to show the heights of the rabbits in our cohort. The observations were (`r paste(myTable$height, sep=',')`). We have four individual rabbits but two share the same height, which will hide one data point since the data will be plotted on top of each other. One way to deal with this is to add a random horizontal scatter to the points using R's *jitter()* function.

```{r, fig.height=2, fig.width=2,  left.mar=TRUE, echo=FALSE}
plot( c(1,1,1,1), myTable$height, ylim=c(0,10), xlim=c(0.8, 1.2),
      ylab="height in inches", xlab="", axes=FALSE, main="No jitter", pch=19 )
box()
axis(2, seq(from=0, to=10, by=2), las=1)


x_locations = jitter( c(1,1,1,1), factor=5 )
plot( x_locations, myTable$height, ylim=c(0,10), xlim=c(0.8, 1.2),
      ylab="height in inches", xlab="", axes=FALSE, main="With jitter", pch=19 )
box()
axis(2, seq(from=0, to=10, by=2), las=1)
```

Plotting individual points this way obscures the connection between a particular subject and its height. If that relationship is important to convey, a **bar plot** would be better choice. The height of each bar is determined by the numerical value of the data. We'll use the *barplot()* function in R. 

```{r, fig.height=2, fig.width=2.5, echo=FALSE}
par(mar=c(5,4,1,1))
barplot( myTable$height, names.arg=rownames(myTable), las=2,
         ylab="height in inches", xlab="", ylim=c(0,10), col="black"  )
box()
```

Histograms summarize continuous data
----------------------------------------------------------------

Once we start gathering a lot of data, it becomes unwieldy and uninformative to look at individual values, either as two-dimensional plots or as a barplot. Here is a set of 1000 observations:

```{r, fig.height=3, fig.width=8, left.mar=TRUE, echo=FALSE }
plot(my_observations,  pch=19, ylab="", xlab="",
     ylim=c(0,100), axes=FALSE, main="Observation 1")
box()
axis(2, seq(from=0, to=100, by=10), las=1)
```

We can see some trends right away: the data tend to lie between 30 and 70, and no value is larger than 90 nor smaller than 10. However, it's not clear where the most common value is. It would be convenient to summarize these data in a simpler format that preserves the important information while reducing clutter

A **histogram** is a helpful method to summarize our data and look for trends. The simplest form of histogram is constructed by dividing the possible values of the observed data into bins, and then counting how many observations fall into each bin. The histogram barplot then shows the count of values in each bin, rather than the raw data itself:


```{r, fig.height=2.5, fig.width=4, med.mar=TRUE, echo=FALSE}
h=hist(my_observations, las=1, xlab="", 
        col="lightgray", main="Obs. 1: histogram (frequency)")
```

```{r echo=FALSE}
freq_break_begin=h$breaks[which(h$counts==max(h$counts))][1]
freq_break_end=h$breaks[which(h$counts==max(h$counts))[1]+1]
```

This histogram is divided into bins of size `r h$breaks[2]-h$breaks[1]`. The bin that spans the range `r freq_break_begin` to `r freq_break_end` has `r max(h$counts)` observations, the most frequent range. We can see that the frequency of values larger or smaller than this range trails off steadily in either direction. 

**Density plots**

An alternate way to display a histogram is a called a *density plot*. In a density plot, the area covered by the histogram bars is constrained to sum to 1. Note that area is the product of bar height and bar width, so short wide bars can cover as much area as tall narrow bars.

```{r, fig.height=2.5, fig.width=4, med.mar=TRUE, echo=FALSE}
h=hist(my_observations, las=1, xlab="", freq=FALSE, ylim=c(0,0.05),
       col="lightgray", main="Obs. 1: histogram (density)")
```

**The sum of the areas of bars in a density plot will always add up to 1**. Plotting the same data with fewer bars will produce a histogram with wider bars, maintaining the total area sum of 1.

```{r, fig.height=2.5, fig.width=4, med.mar=TRUE, echo=FALSE}
h=hist(my_observations, las=1, xlab="", freq=FALSE, ylim=c(0,0.05),
       col="lightgray", main="Obs. 1: histogram (density)", breaks=5)
```

Observing a summary of all of the values on a density plot allows us to make some observations about frequently observations take a given value. This gives information about the **distribution** of values that we observed. We will explore distributions in more detail in Module 4.

**How many peaks?**

Histograms can also make it clear whether there is a single peak of most frequent values, or more than one such peak. However, careful study of the data may be required to make this clear. A new dataset of 2000 observations makes this point:

```{r, fig.height=2.5, fig.width=8, left.mar=TRUE, echo=FALSE, left.mar=TRUE }
plot(sample(my_bivariate),  pch=19, ylab="", xlab="",
     ylim=c(0,100), axes=FALSE, main="Observation 2")
box()
axis(2, seq(from=0, to=100, by=10), las=1)
```

```{r, fig.height=2, fig.width=4.5, med.mar=TRUE, echo=FALSE, left.mar=TRUE }
h=hist(my_bivariate, las=1, xlab="", main="Observation 2: histogram", ylim=c(0,400), col="lightgray" )
```

In this second observation, a histogram makes it clear that there are two peaks in the frequencies, one around 50 and one around 65. The shape of the distribution suggests something about the process that generated the data. Statisticians refer to the most frequent value in a data set as the *mode* of the distribution, and a distribution of observations with two peaks is called a *bimodal distribtion*.

**Bimodal distributions in mouse blood**

In 2013, the Balmain lab performed a Complete Blood Count (CBC) analysis on whole blood taken from `r length(MCV)` healthy eight week old mice. The CBC panel is a standard measure of macroscopic blood phenotypes such as the relative frequency of red blood cells in the total population. The Mean Corpuscular Volume is an estimate of the average red blood cell size, measured in fL.

Plotting the measured values using a histogram gives information about the distribution of values. Here I present two different histograms from the same data, one with fewer bars and another with more bars:

```{r, fig.height=2.5, fig.width=3, left.mar=TRUE, echo=FALSE, left.mar=TRUE }
hist(MCV, breaks=5, main="MCV, wider bins", col="lightgray", xlim=c(40,80), ylim=c(0,80), las=1 )
hist(MCV, breaks=20, main="MCV, narrower bins", col="lightgray", xlim=c(40,80), ylim=c(0,40), las=1 )
```

In the plot on the left, each bar represents a wider bin of data. This plot suggests there is a single peak of frequency around 45 fL. The plot on the right shows the same data plotted using smaller bins, and suggests that there may actually be two distinct peaks of MCV levels in this population, one around 45 fL and one around 55 fL. The mice in this study were a genetically heterogeneous mix of two parental strains, and it turns out that control of MCV levels was strongly associated with the individual mouse's genotype at one locus in the genome.

R code examples
=======================

> Convert strings to factors

We will collect six measurements for eye color, store them to a vector called eye_color, and then convert the the vector to a factor

```{r}
eye_color = c("brown", "blue", "blue", "green","blue","grey")
eye_color
f_eye_color = factor(eye_color)
f_eye_color
```

Note that after using the *factor()* function, printing out the variable *f_eye_color* returns both the strings and a list of the **Levels**, or unique names, that make up the factor.

> Scatter plot a vector of data

The simplest plot of a single vector of size *N* assigns increasing values 1..*N* for the X axis and plots the vector on the Y axis:

```{r fig.height=3, fig.width=3, small.mar=TRUE}
data=c(1,3,5,2,7)
plot( data )
```

This plot can be modified in many ways. Some basic parameters:

* Figure header: *main* (words in quotes)
* X label: *xlab* (words in quotes)
* Y label: *ylab* (words in quotes)
* Point style: pch (various numbers or characters. Some useful ones are 1, 2, 3, 19, and '.' (the period). If you don't pass a value, the value is set to 1 automatically. If you pass a single value, all points have the same point style. Alternately, you can pass a vector of values and set a point styles for each point individually, which is useful if you want to differentiate different kinds of measurements
* Point color: col (word in quotes or RGB code in quotes)

```{r fig.height=3, fig.width=3, med.mar=TRUE}
data=c(1,3,5,2,7)
plot( data, pch=19, col="red", main="Example header", xlab="X label", ylab="Y label" )
```

A full tutorial on making plots is beyond the scope of this module. Many people find the *ggplot2* package to be very helpful in eliminating many of the technical details of making good-looking plots.


> Plotting Bar Plots

Pass a vector of numbers to the bar plot in the order in which the values should be plotted

```{r fig.height=3, fig.width=5, small.mar=TRUE}
barplot( c(1,3,5,2,7 ) )
```

Useful parameters:

* labels: names.arg (vector of words in quotes)
* Figure header: *main* (words in quotes)
* X label: *xlab* (words in quotes)
* Y label: *ylab* (words in quotes)
* Bar color: *col* (word in quotes or RGB code in quotes)
* Add second bar plot on top of existing plot: *add*=TRUE

```{r fig.height=3, fig.width=5, small.mar=TRUE}
barplot( c(1,3,5,2,7 ), col="black", names.arg=c("one", "two", "three", "four", "five") )
```

Barplots can be layered on top of each other by plotting a second bar plot with the add parameter

```{r fig.height=3, fig.width=5, small.mar=TRUE}
barplot( c(1,3,5,2,7 ), col="black", names.arg=c("one", "two", "three", "four", "five") )
barplot( c(0,2,4,1,5 ), col="cornflowerblue", add=TRUE )
```



> Plotting Histograms

Histograms can be plotted with the hist() function. By default a frequency histogram is plotted:

```{r fig.height=3, fig.width=5}
hist(my_observations)
```

Several useful parameters:

* Main title: *main* (words in quotes)
* X label: *xlab* (words in quotes)
* Y label: *ylab* (words in quotes)
* number of bins: *breaks* (number)
* bin color: *col* (color name or RBG code in quotes)
* plot a density histogram: pass freq=FALSE

```{r fig.height=3, fig.width=5}
hist(my_observations, freq=FALSE, col="cornflowerblue", breaks=20, 
     main="My observations", xlab="days until relapse")
```
